---
editor_options: 
  markdown: 
    wrap: 72
---

# Exploratory Data Analysis (EDA)

In this Chapter we start into the main strength of R: Data Analysis.
After cleaning our data we are ready to analyze it, but it is always
recommend to get an overview of our data with exploratory data analysis.
This yields the advantage that we can detect first trends and potential
problems with our data. Furthermore, we will get a better understand of
the Data Generating Process and how data is generated.

Let us tart with loading packages and you will see, that we are loading
a lot of packages due to the fact, that R is a statistical analysis
software and therefore it is only natural that Statisticians and Data
Scientists, who use R, are also writing packages to (1) make their own
life easier, but also (2) further developing R.

```{r loading packages eda}
pacman::p_load("summarytools", "smartEDA", "skimr", 
               "naniar", "Hmisc", "gtsummary", "dlookr",
               "DataExplorer", "psych", "ggplot2")
```

## Descriptive Statistics Fundamentals

The absolute Basics of EDA are metrics like mean, median and standard
deviation, but efore we look at them it is crucial to understand, where
these metrics come from. And to understand that we have to understand
how Data is generated.

### Random Variable

We are interested in Data! And Data is nothing more than a bunch of
**Random Variables.** Technically speaking, a Random Variable is a
numerical outcome of a random process or experiment. Informally, let us
say you want to collect Data about your family, just image all of them
and now randomly pick some information about them, their height, their
age, gender, glasses or not, cooking abilities, anything, and now you
just need to find a system of assigning a number (e.g. height in cm, age
in years,...) and et violà you collected a random variable.

Let us say we collect the age of my mum, and if she wears glasses or not
(1 = yes, 0 = No)

```{r random variable}
#creating a random variable
age_mum <- 47
glasses_mum <- 1

#Print it
print(age_mum)
print(glasses_mum)
```

Well, my mum is 47 and she wears glasses. Both information, numerically
expressed in random variables (age_mum, and glasses_mum).

### Probability Distributions

This data about my mom gives us information about one person, but
working with quantitative methods means, to (guess what?) collect data
quantitatively. Thus, we are mostly interested in a so-called
**population.**

In statistics, a **population** refers to the entire set of individuals,
items, or data points that are of interest in a particular study. It
represents the complete group about which you want to draw conclusions
or make inferences.

In my example, the population I would be interested in could not only be
my mum, but all mums living in Germany. I want to know their heights and
if they wear glasses or not.

To get all the heights and if they wearing glasses of all mums in
Germany would be a lot. According to the public statistic there are 20
million mothers in Germany. But that is no problem, we can take smaller
subset of the population, selected for analysis. This is called a
**sample**. Let us simulate 1000 mothers, their heights and if they wear
glasses and simulate it in R, put it in a data frame and visualize it:

```{r simulating sample}

#Simulating sample
sample_mum <- data.frame(
  name = rep("mum", 1000),
  height = rnorm(1000, mean = 165.8, sd = 9),
  glasses = as.factor(rbinom(1000, 1, 0.33))
)

#print it
head(sample_mum)

#Making a histogram for the heights
ggplot(sample_mum, aes(x = height)) +
  geom_histogram(binwidth = 2, color = "white", fill = "#FC4E2A") +
  labs(
    x = "Height (in cm)",
    y = "Count"
  ) +
  theme_bw()

#Making a bar plot for glasses
ggplot(sample_mum, aes(x = glasses)) +
  geom_bar(fill = "#FC4E2A", width = 0.5) +
  labs(
    x = "",
    y = ""
  ) +
  scale_x_discrete(labels = c("0" = "No Glasses", "1" = "Wearing Glasses")) +
  theme_bw()

```

Random variable and Sampling PDFs and CDFs Median, Mean, Mode Variance
and Standard Deviation Quantiles & IQR

summarytools: Showcase descr() to display summary statistics with
additional formatting. skimr: Use skim() to summarize a dataset and its
variables. Hmisc: Show functions like describe() for comprehensive
descriptive statistics.

## Summarizing Categorical Data

Frequency Tables & Proportional Tables Package Focus: gtsummary:
Demonstrate how to generate detailed summary tables for categorical data
with tbl_summary(). smartEDA: Show how to use smartEDA() to generate
statistics and visualizations for categorical variables. dlookr:
Highlight its automatic analysis and visual presentation of categorical
variables.

## Correlation

Cross-Tabulations Correlational Analysis

## Conclusion

Key Take aways

## Exercises
