# Data Analysis

In this chapter you I introduce you to the basic ideas of statistical analysis and hypothesis testing. For me, it is important that you get an intuition about what is going on rather than terrorizing you with complicated math. Although, I cannot and will not leave out central formulas, you will be fine with the maths. There extensively commented and in the end of the day not that hard to understand. You will notice that the programming in this chapter is quite easy compared to before. In the end, you only need formulas, who do everything you need to analyse data, that is also a strength of R. Please concentrate more on the concepts and get an idea what is going on. The goal of this chapter is that you get to know linear regression and how to check if the model fits and the results are robust and significant.

```{r loading packages}
pacman::p_load("tidyverse", "ggpubr","gapminder", "sjPlot", "kableExtra", 
               "GGally", "car", "margins", "plotly")
```

## Linear Regression

### Terminology

The classical (bivariate) linear regression can be expressed in the following equation (systematic component):

$$
Y_i = \beta_0 + \beta_1X_i + e_i 
$$

where

-   the index *i* runs over the observations (Respondents, Countries,...), *i* = 1,...,*n*

-   $Y_i$ is the dependent variable, the variable we want to explain

-   $X_i$ is the independent variable or explanatory variable.

-   $\beta_0$ is the *intercept* of the regression line

-   $\beta_1$ is the *slope* of the regression line

-   $\epsilon_i$ is the *error term*, thus how our observed data differs from actual population data (e.g. Measurement Error).

### Estimating the Ordinary Least Squares Estimator

To get the idea of linear regression, let us look at an example. To do so, let us simulate some data and plot it.You now should have the data frame `df` in your environment. It contains a variable X, which is our independent variable. Y is also included, which is your dependent variable. You want to explain Y with your X. Let us plot the variables with a scatterplot:

```{r simulating data, echo=FALSE}

set.seed(123) # For reproducibility

n <- 30 

x <- runif(n) * 10 

categorical_variable <- factor(sample(c(0, 1), n, replace = TRUE))

y <- 0.8 + 1.6 * x + rnorm(n, 0, 3)

df <- data.frame(x,y, categorical_variable)

#Simulate further data

X_quadratic <- X <- runif(50, min = -5, max = 5)
u <- rnorm(50, sd = 1)  

#True relation
Y_quadratic <- X^2 + 2 * X + u

#Making a data frame out of it
df2 <- data.frame(X_quadratic, Y_quadratic)

#Simulating time data 

# set seed
set.seed(123)

# generate a date vector
date <- seq(as.Date("1960/1/1"), as.Date("2020/1/1"), "years")

# initialize the employment vector
y_time <- c(5000, rep(NA, length(date)-1))

# generate time series observations with random influences
for (i in 2:length(date)) {
  
    y_time[i] <- -50 + 0.98 * y_time[i-1] + rnorm(n = 1, sd = 200)
}

# creating DataFrame 
df_time_series <- data.frame(y_time, date)

###Defining a function

#Creating Function for Kable Table 
table_ovb <- function(model1, model2) {
  # Determine the maximum number of coefficients between the two models
  max_coef <- max(length(model1$coefficients), length(model2$coefficients))
  
  # Initialize placeholders for coefficients
  place_holder1 <- rep(NA, max_coef)
  place_holder2 <- rep(NA, max_coef)
  
  # Replace the placeholders with coefficients from the models
  place_holder1[1:length(model1$coefficients)] <- model1$coefficients
  place_holder2[1:length(model2$coefficients)] <- model2$coefficients
  
  # Create a data frame with coefficients from both models
  dt <- data.frame(place_holder1,place_holder2) 
  
  # Set row and column names
  colnames(dt) <- c("Model without Temperature", "Model with Temperature")
  rownames(dt) <- c("Intercept", "Ice Cream Sales", "Temperature")
  
  # Display the table
  
dt %>%
  kbl() %>%
  kable_styling()
  
}
```

```{r plotting x and y in df}
ggplot(df, aes(x, y)) + 
  geom_point() +
  theme_bw() + 
  scale_x_continuous(breaks = seq(0, 10, by = 1)) +
  scale_y_continuous(breaks = seq(0, 20, by = 2))
```

We can see that there has to be some relationship between both those variables: The higher x gets the higher y gets. Linear regression can help us investigate the relationship. We just have to take the formula and estimated $\beta_0$ and $\beta_1$ . To do so, we estimate the **ordinary least square (OLS) estimator**. To understand what the OLS estimator does, look at the scatter plot again: **The OLS estimator fits a line through all the dots, that minimizes the distance to the dots as much as possible**. Afterward, we only extract the intercept $\beta_0$ (that is the point, where the line crosses the y-axis), and $\beta_1$ (that is the slope of the line). However, since these are estimated values for our model we have to call them by convention $\hat{\beta_0}$ and $\hat{\beta_1}$ . We will denote them in the following as such.

#### Visualization

The visual estimation is good to get an intuition with the data, but we will see later, that it does not work with multiple independent variables. Further, it does not give much information, at least not as much as we would like to have.

```{r fitting an ols estimator line visually}
ggplot(df, aes(x, y)) + 
  geom_point() +
  theme_bw() + 
  scale_x_continuous(breaks = seq(0, 10, by = 1)) +
  scale_y_continuous(breaks = seq(0, 30, by = 5)) + 
  geom_smooth(method = "lm", se = FALSE) + 
  geom_segment(aes(x = x, y = y, xend = x, yend = predict(lm(y ~ x, data = df))), linewidth = 0.5) 
```

#### Calculation per Hand

We could also just calculate $\hat{\beta_0}$ and $\hat{\beta_1}$ by hand. The formula for both are as follows:

$$
\hat{\beta_1} = \frac{\sum_{i=1}^n(x_i - \bar{x})(y_i - \bar{y})}{\sum_{i=1}^n(x_i - \bar{x})^2}
$$

where $x_i$ is the answer of respondent *i* for our independent variable x, $\bar{x}$ is the average answer of the respondents. Those two values are subtracted, thus the deviation from the mean is calculated. The same goes with our dependent variable y, where $y_i$ is the answer of respondent *i*, and $\bar{y}$ is the average answer of all respondents. his is done for every respondent *i*, thus *n* - times. This is displayed with the sum symbol. We just calculated the so-called **covariance.**

The **covariance** is then divided by the squared deviation to the average answer of our independent variable x. This will get us the estimated coefficient for our model $\hat{\beta_1}$ . Sounds complicated but lets do it in R:

```{r calculating ols by hand}
#Let us first get the covariance
cov <- sum((df$x - mean(df$x)) * (df$y - mean(df$y)))

#Now we get the variance of x 
x_sq <- sum((df$x - mean(df$x))^2)

x_sq

# We just have to divide them 
slope <- cov/x_sq 

#printing it
print(slope)
```

To get the intercept $\hat{\beta_0}$ we have to take the result of the slope and implement it into this formula:

$$
\hat{\beta_0} = \bar{y} - \hat{\beta_1}*\bar{x}  
$$

We multiply $\hat{\beta_1}$ with the average of our independent variable X. The result is subtracted from the average of our dependent variable y.

```{r calculating intercept by hand}
#calculating the intercept
beta_0 <- mean(df$y) - (slope * mean(df$x))

#printing it
beta_0
```

Now we estimated our parameters and can display the for our model by simply putting it into the systematic component:

$$
Y_i = 1.68 + 1.54 * X_i
$$

#### Automated Calculation

This procedure by hand is way to time-wasting. R has a built in function to calculate the parameter for us called `lm()` :

```{r calculating linear regression automated}
#running a linear regression
model1 <- lm(y ~ x, 
             data = df) 

#Printing a summary of the model results
summary(model1)
```

When we check the results, we see that we did everything right and get the exact same values. The interpretation of the coefficient is with a one-unit increase in the independent variable X, the dependent increases on average about 0.68 units, holding all else constant. But you noticed that you get more information on the model, than just the coefficients. We will get to that later.

### Predictions with Linear Regression

We could calculated predictions based on our OLS calculations, reconsider the systematic component we calculated:

$$
\hat{y_i} = 1.68 + 1.54x_i 
$$

We just have to put in x-values and according to our model we would get a prediction of the respondents y-value $\hat{y_i}$. We can do that for all x-values in our dataset and get a column with all the predicted values, let us call it (`y_hat`) :

```{r calculating predictions}
#First, we calculate the predictions for y
df$y_hat <- 1.6821 + 1.5394*df$x 

#We could also do it automatically via the predict() function
df$auto_y_hat <- predict(model1)

#Checking it 
head(df)
```

## Model Fit

Now the linear regression has a lot of assumptions, it is not like we can run the model every time how we want. Since it is a model, it makes assumptions and instead of just assuming them to be right, we can test them. To techniques to test them are called **Measures of Fit**. Because they test how much our data fits the data. Let us have a look at the assumptions and how we can test them:

### Measures of Fit

#### Residuals

1.) Calculating Residuals:

$$
Residuals = y_i - \hat{y_i} = y_i - (\hat{\beta_0} - \hat{\beta_1}x_i) 
$$

where

-   $y_i$ = our actual observed values of our dependent variable (`df$y`)

-   $\hat{y_i}$ = are our predicted values based on our OLS estimator

Reconsider the graph at 2.2.1, the residuals are basically the red lines, thus the distance from the line to the point. We can calculate the residuals for our graph:

```{r calculating residuals}
#We get the Residuals by subtracting our actual y from y_hat
df$residuals <- df$y - df$y_hat 

#cheking it
head(df) 

#We could have done that automatically with R as well !  
df$residuals_auto <- residuals(model1)

#Checking it
head(df)
```

Let us have a look at a so-called **Residual Plot**: On the x-axis you plot the **fitted values**, thus our `y_hat`. On the y-axis you plot the residuals, thus $y-\hat{y}$. Then you plot a horizontal line at y = 0. All dots on those lines show us the values correctly predicted by our model.

```{r residual plot}
ggplot(df, aes(x, residuals_auto)) + 
  geom_point() + 
  geom_hline(yintercept = 0) +
  scale_y_continuous("Residuals", seq(-6, 6, 1), 
                     limits = c(-6, 6)) +
  scale_x_continuous("Fitted Values", seq(0, 10, 1), 
                     limits = c(0, 10)) +
  theme_bw()
```

#### Homoskedasticity and Heteroskedasticity

One assumption of linear regression is that the variance of the error term is not correlated with our independent variable. Well, that is quite technocratic and means basically, that the residuals are distributed equally over the independent variables. Let us plot it to get a visual intuition:

```{r homo and hetero plot, message=FALSE, warning=FALSE}
#setting seed for reproduciability
set.seed(123)

# Generate some data
x <- runif(150, 0.05, 1)
e <- rnorm(150, 0, 0.5)

#homoskedastic data 
y_homo <- 2 * x + e 
#heteroskedastic data 
y_hetero <- 2 * x + e*x^2 
#making a data frame with both data
df_homo_hetero <- data.frame(x, y_homo, y_hetero)

# Scatterplot with homoscedasticity
homoskedastic_plot <- ggplot(df_homo_hetero, aes(x = x, y = y_homo)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +# Add linear regression line
  scale_y_continuous("Y", seq(-0.5, 3.5, 0.5), limits = c(-0.5, 3.5)) +
  labs(title = "Homoskedastic Plot") +
  theme_minimal()

# Scatterplot with heteroscedasticity
heteroskedastic_plot <- ggplot(df_homo_hetero, 
                               aes(x = x, y = y_hetero)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +#Add linear regression line
  labs(title = "Heteroskedastic Plot") +
  scale_y_continuous("Y", seq(-0.5, 3.5, 0.5), limits = c(-0.5, 3.5)) +
  theme_minimal()

# Combine plots using facet_wrap
facet_plots <- ggarrange(homoskedastic_plot, heteroskedastic_plot, nrow = 1)

# Print the combined plots
print(facet_plots)
```

In the left plot, you see the homoskedastic data. The dots are equally and constantly distributed around the fitted line. However, the right plot shows that the more the independent variable **x** increases, the more the observations are increasing. The dots are not constantly distributed over the line. In the case, that the data is heteroskedastic, then this is a problem. You could try to transform the independent variable by taking the logarithm (We will look into that later). You could also use so-called heteroskedastic regression, but this an advanced model.

### TSS, ESS and $R^2$

Now that we have the Residuals, we can calculate the Total Sum of Square (TSS), the Explained Sum of Squared ESS, and $R^2$:

-   TSS (Variation in the DV): $TSS =\sum(y_i - \bar{y})^2$ , we just subtract our actual values (`df$y`) from its mean and square it to avoid negative numbers. This gives us the total variation of our dependent variable.

-   ESS (Variation we explain in the DV): $ESS = \sum(\hat{y_i} - \bar{y})^2$ , now we use our predicted values (`df$y_hat`) instead of our actual values. That gives us the variation in the dependent variable, we can explain with our model.

-   $R^2$ (The Variation we can predict from our model): $R^2 = \frac{ESS}{TSS}$ , well to get the proportion we just divide the variation we can explain from our DV from the actual variation through the total variation in the DV. If these two values are the same, thus our model predicts all the variation in our dependent variable and this $R^2$ is 1. If our model could not explain anything the variation would be 0, since the values of both cannot be negative. Let us calculate them:

```{r calculating r2 by hand}
#total sum of squares
tss <- sum((df$y - mean(df$y))^2)
#explained sum of squares
ess <- sum((df$y_hat - mean(df$y))^2)
#caculating r squared
r_squared <- ess/tss

#Printing it
r_squared

#Summarizing it
summary(model1)$r.squared
```

### Influential Outliers

Outliers are extremely deviating values, which can impact our analysis and bias it. Therefore, we have to check, if our data contains such values. But first let us see how they can impact our data:

```{r plotting bias through outliers}
#set seed 
set.seed(069)

#generate fake data with outlier 
x1 <- sort(runif(10, min = 30, max = 70))
y1 <- rnorm(10 , mean = 200, sd = 50)
y1[9] <- 2000
data_outlier <- data.frame(x1, y1)

#Model with Outlier 
model_outlier <- lm(y1 ~ x1) 

#Model without Outlier
model_without_outlier <- lm(y1[-9] ~ x1[-9]) 

#Plotting the Data 

# Scatter plot with points
ggplot(data_outlier, aes(x = x1, y = y1)) +
  geom_point(shape = 20, size = 3) +
  # Regression line for the model with outlier
  geom_abline(aes(slope = model_outlier$coefficients[2], intercept =
              model_outlier$coefficients[1], 
              color = "Model with Outlier"), linewidth = 0.75, show.legend = TRUE) +
  # Regression line for the model without outlier
  geom_abline(aes(slope = model_without_outlier$coefficients[2], 
              intercept = model_without_outlier$coefficients[1], 
              color = "Model without Outlier"), linewidth = 0.75, 
              show.legend = TRUE) +
  xlab("Independent Variable") +
  # Adding legend
  theme_classic() + 
  theme(legend.position = c(0.15,0.9), 
        legend.title = element_blank()) 

```

We see that this one observation completely biases our sample. But how do we find out, which observation is an influential outlier? There is a metric called **Cook's Distance**, we can use. Let us do it and plot it in R.

```{r plotting cooks distance}
#Cooks Distance can be calculated with a built-in function
data_outlier$cooks_distance <- cooks.distance(model_outlier) 

#Plotting it
ggplot(data_outlier, aes(x = x1, y = cooks_distance)) + 
  geom_point(colour = "darkgreen", size = 3, alpha = 0.5) + 
  labs(y = "Cook's Distance", x = "Independent Variables") + 
  geom_hline(yintercept = 1, linetype = "dashed") + 
  theme_bw()
```

We can clearly see that **Cook's Distance** detected the outlier. The rule is that values with a cooks distance bigger than 1 have to be eliminated. You can do that with the `filter()` function and run the model afterward again without the outliers.

### Functional Form

Linear regression is a mathematical model. Therefore it is based on assumptions. But we should not just assume them, we should test them! One assumption is that linearity is assumed between X and Y. But that can be problematic consider following example:

```{r plotting linear fit through quadratic form}
# estimate a simple regression model 
model_simple <- lm(Y_quadratic ~ X_quadratic, data = df2)

# Summarize it
summary(model_simple)

#Plot it
ggplot(df2, aes(x = X_quadratic, y = Y_quadratic)) + 
  geom_point(shape = 20, size = 3) + 
  geom_smooth(method = "lm", se = FALSE) + 
  theme_bw()
```

As you can see something look wrong. There seems to be a correlation between the two variables, but it does not seem linear. In such a case it does make sense to square the independent variable and run the regression again:

```{r plotting quadratic fit with quadratic form}
# estimate a simple regression model 
model_quadratic <- lm(Y_quadratic ~ X_quadratic^2, data = df2)

#Summarize it 
summary(model_quadratic)

#Plot it
ggplot(df2, aes(x = X_quadratic, y = Y_quadratic)) + 
  geom_point(shape = 20, size = 3) + 
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), 
              color = "red", 
              se = FALSE,) + 
  scale_x_continuous("X", breaks = seq(-5,5,1), limits = c(-5,5)) +
  ylab("Y") +
  theme_bw()
```

Well that looks better and is the proper way to deal with quadratic relationships in linear regression. Well, data can take on not only a quadratic form, it could also take on a form of a square-root function. I will show the most classical example of such a functional form. The `gapminder` data is loaded. It contains data about the average life expectancy (`lifeExp`)and the GDP per capita (`gdpPercap`) of countries in different years. Let us look if the GDP per Capita is correlated with Life Expectancy:

```{r unlogged fit, warning=FALSE}
#checking the data
head(gapminder)

#Plotting it
ggplot(gapminder, aes(gdpPercap, lifeExp)) + 
  geom_point() + 
  geom_smooth(method = "loess", se = FALSE) + 
  scale_y_continuous("Life Expectancy", seq(30, 80, 10), 
                     limits = c(30, 80)) + 
  theme_bw()
```

Well, that looks terrible. What can we do? I already mentioned that the fitted line looks like a square-root function ($y = \beta{\sqrt{x}}$ ). When you take the logarithm of square root, you neutralize the square root and only x remains $\log{(\sqrt{x})} = x$. When you do that the functional form changes to $y = \beta{x}$. Well, that is exactly the systematic component we are after:

```{r logged fit, warning=FALSE}
ggplot(gapminder, aes(log(gdpPercap), lifeExp)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) + 
  scale_y_continuous("Life Expectancy", seq(30, 80, 10), 
                     limits = c(30, 80)) + 
  xlab("GDP per Capita") +
  theme_bw()
```

This looks more like what we want to achieve. What can we see in that plot? When we run a model we want to take the logarithm of the independent variable, when we expect the following: If the most observations of a variable are low, but some observations are extremely high such functional forms can occur. In our example, most of the countries have a low GDP per Capita, but some countries such as the Western European countries or the USA have such a a high level of GDP per Capita, they change the functional form of the fitted line. These are influential outliers, but too much to delete them. It could bias the representativeness of the sample, therefore we can deal with them by taking the logarithm.

### Independent Observation

Another assumption of the linear regression model is the independent, identically distributed (i.i.d) assumption. That sounds complicated but it really is not. Consider following plot:

```{r time trend}
ggplot(df_time_series, aes(date, y_time)) + 
  geom_line() +
  ylab("Y") +
  xlab("Year") +
  theme_bw()  
```

If you look at the plot, can we assume that the observation Year = 2000 is independent from the Years before? No, the observation in the years are correlated to each other, thus the assumption is violated. This is basically the huge problem of working with longitudinal data (time-series cross-sectional or panel). If you face such problems there are plenty of other methods to use: Interrupted time series, Difference-in-Difference Designs, Panel-Matching, Fixed-Effects Models etc.

## Hypothesis Testing in R

We are not only interested if our model fits or not. We want to investigate real world phenomena. Well, and if our model does not fit well, we have to make adjustments so it does. What comes next is that we want to know more about the world. Researcher do so by formulating **hypotheses**. These are nothing more than assumptions you make theoretically about the world. Let us say you think you assume that in our world education has an impact on income. This would be your **Alternative Hypothesis** $H_A$ . What you know want to do is to test it against the so-called **Null Hypothesis** $H_0$. That is nothing more than the opposite of our alternative hypothesis, thus that education has no impact on income. Let us formulate both to have an overview:

$H_0$ = Education does not impact Income.

$H_A$ = The more educated a person is, the higher the income.

The advantage of this approach is obvious, one of those will be true. Therefore statistical testing is needed. But before introducing it to you, we have to decide how we want to test our alternative hypothesis. More specifically, how do we want to measure our variables. For our example, we decide to conduct a survey and to get data by asking a random sample of 1000 people over 18 living in Germany (*N*=1000). We decide to measure our independent variable Education, by asking the respondents about the years they invested in their education. To get their income, you just ask them about it to fill it in.

Can you think about possible critics about our data collection strategy?

### Standard Error

#### Root Mean Square Error (RMSE)

Before moving on to Standard Errors, I will introduce another metric, the root mean square error. It is the average difference between the actual values $y_i$ and our predicted values $\hat{y_i}$. To calculate it, we square the residuals, to get only positive values. Then we take the mean, and lastly we take the square root.

```{r RSME}
#Getting the the sum of squared residuals (SSR)
SSR <- df$y_hat^2

#Calculating the mean of the squared residuals 
mean_SSR <- mean(SSR) 

#Calculatin the RSME 
rsme <- sqrt(mean_SSR)

#Printing it 
print(rsme)
```

The rule of thumb is that the lower the RSME, the better. It is a non-standardized goodness of fit measure. Its counterpart is the R-squared measure, which is a standardized measure. You can use both, and should use both to get a metric about how close the predicted values are distributed around the actual values.

#### Standard Error of the Estimate

The standard error of a coefficient estimate is the metric directly presented next to the coefficient in the regression output. It is the square root of the variance of the regression coefficient.

```{r se}
#calculating standard error by hand
se <- SSR/(nrow(df) - 2 * (sum(df$x - mean(x))))

#Printing it
print(se)
```

### T-Value or T-Statistic

The formula for calculating the t-value is simple:

$$
t_i = \frac{\beta_i}{SE(\beta_i)}
$$

where $\beta_i$ is the coefficient calculated by the linear regression, and $SE(\beta_i)$ is the standard error. Let us calculate it manually by hand for our example:

```{r t-value by hand}
#t value by hand
t_value_intercept <- -0.32773/0.30271 
t_value_x         <- 0.67949/0.04813 

#printing it
print(t_value_intercept) #-1.082653
print(t_value_x) #-14.11781
```

Well, that is the t-value on its own does not tell us about statistical significance. It is used to calculate the **p-value**, which tells us about the significance of the value. But before we calculate it, we have to understands some key concepts before.

-   **Degrees of Freedom** is the first concept. Let us say, we have a sample of me and my sister (*N* = 2). We collected the numbers of books each of us has, and calculated a mean of 30. I have 20 books. How many books does my sister have? Obviously, she has 40, if we have a mean of 30. Given my value, and the mean of the sample, the value of my sister had to be 20. That is what the degrees of freedom tells us. Given the mean of a sample, the values, which can freely be chosen. Or to put it more technically, the maximum number of logically independent values.

-   The rule is that **the larger the sample, the higher the degrees of freedom** and **the lower the sample, the lower the degrees of freedom.**

-   **T-distributions** are the distribution, we will use to calculate the p-value. I will talk about that in detail in a minute. But they are connected to the degrees of freedom, because degrees of freedom determine the tail behavior and shape of the curve until the point, where the t-distribution looks like a standard normal distribution. Let us visualize that:

    ```{r interactive t-statistics plot}
    # Generate data
    x <- seq(-5, 5, length.out = 100)

    # Calculate densities
    densities <- data.frame(
      x = rep(x, 4),
      density = c(dt(x, df = 1), dt(x, df = 2), dt(x, df = 10), dnorm(x, mean = 0, sd = 1)),
      distribution = rep(c("t(df=1)", "t(df=2)", "t(df=10)", "Normal"), each = 100)
    )

    densities$distribution <- factor(densities$distribution, 
                                     levels = c("Normal", 
                                                "t(df=10)", 
                                                "t(df=2)", 
                                                "t(df=1)"))

    # Plot
    plotly::ggplotly(ggplot(densities, aes(x = x, y = density, color = distribution)) +
      geom_line() +
      theme_minimal() +
      labs(x = "x", y = "Density", 
           title = "t-distributions with different degrees of freedom") + 
      scale_color_manual(values = c("black", "red", "green", "blue")) + 
      scale_x_continuous("X", seq(-5,5,1), limits = c(-5,5))) 
    ```

-   Alright, before we find out, how to determine if a value allows us to reject the null hypothesis, we have to determine a so-called **critical value**. This is no math or anything, it is a probability we decide about. More precisely, the probability that our t-value and thus our coefficient occured by chance alone. If we say the probability that it occured by chance alone is 10%, then this is our critical value. The rule of thumb is, that a probability lower than 5% that the coefficient occured by chance alone indicates statistical significance. Keep in mind, that the critical value can vary depending on the sample size, the degrees of freedom, the field you are working in etc.

```{r t-static identification visually}
#setting seed 
set.seed(42) 

# Generate data
x <- seq(-5, 5, length.out = 100)
t_density <- function(x) dt(x, df = 28)

# Calculate densities
t_value_data <- data.frame(
  x = rep(x, 1),
  density = dt(x, df = 28),
  distribution = rep("t(df=28)", 100)
)

# Plot
plotly:: ggplotly(ggplot(t_value_data, aes(x = x, y = density)) + 
  geom_line(lineend = "round") + 
  stat_function(fun = t_density, geom = "area", fill = "gray", 
                alpha = 0.75, xlim = c(-5, -1.701), n = 10000) +
  stat_function(fun = t_density, geom = "area", fill = "gray", 
                alpha = 0.75, xlim = c(5, 1.701), n = 10000) +
  geom_vline(xintercept = -1.701, linetype = "dashed", 
             colour = "red") +
  geom_vline(xintercept = 1.701, linetype = "dashed", 
             colour = "red") + 
  ggtitle("t-distribution with 28 df", subtitle = "The pink area marks the interval of significant values on a 95% level") +
  geom_segment(x = -1.082653, 
               xend = -1.082653, 
               yend = dt(-1.082653, df = 28), 
               y = -1, 
               color = "pink", 
               linetype = "dashed", 
               linewidth = 0.2) +
    annotate("point", x = -1.082653, y = dt(-1.082653, df = 28), 
             color = "pink") +
  scale_x_continuous("X", seq(-5,5,1), limits = c(-5,5)) +
  theme_classic() +
    theme(legend.position = "none")
)
```

-   As we can see, the blue line representing the value of the intercept is not in the area it would have to be, for us to reject the null hypothesis. However, the t-value of our coefficient from variable is with about 14 far away from the threshold, so we can reject the null hypothesis. This is how the t-value works. The problem is, that the threshold varies with the degrees of freedom, and you will not have the threshold value in your head for every degree of freedom. You could look it up every time or you use **p-values**, which directly tell you the probability of the coefficient occuring by chance alone.

### p-values

-   The p-value is a statistics that shows us the probability that a statistical measure (in our example 0) is greater/less than an observed value (in our example, the estimated coefficient). The null hypothesis would tell us that our independent variable X has no impact on Y. Statistically speaking, that would mean that our coefficient has a high probability to be zero, because zero means no effect, thus we cannot reject the null hypothesis. But if the probability that our estimated coefficient is 0 is low, we may reject the null hypothesis and would found an effect.

-   Before showing you two ways to find the p-value, we have to determine a **critical value**. This is no math or anything, it is a probability we decide about. The rule of thumb is that if the p-value is smaller than 5%, we can reject the null hypothesis. However, depending on various factors, it could also be different, that depends on your data, sample size, degrees of freedom etc.

-   For our example, we follow the rule of thumb and set the significance level to 0.05, thus if the coefficient has a smaller probability (p-value \< 0.05) than 5% to be zero we can reject the null hypothesis, otherwise we cannot reject it.

-   Since the math is complicated and thus not help to understand the intuition, I directly show you how to calculate the p-value by hand:

```{r p-value by hand }
#calculating the p values by hand
p_value_1 <- 2 * pt(-abs(t_value_intercept), 28)
p_value_2 <- 2 * pt(-abs(t_value_x), 28)

#printing it
print(p_value_1) 
print(p_value_2)
```

We get the same values as in our regression, and we decided before that our critical value (denoted as $\alpha$) should be less than 0.05 to reject the null hypothesis. Therefore we can reject the null hypothesis for the coefficient of our variable x. Since the coefficient is positive, the interpretation would be that x has on average a positive effect on y, since the probability that the value is different than 0 is lower than the critical value of 5%.

### Confidence Interval

The last way of testing hypothesis are confidence intervals. I recommend to use them, since they are intuitive and easier to interpret than p-values. To understand confidence intervals, we must remember the difference between a **population** and a **sample.** The population are all people we want to infer to, for example in an election, the population are all citizens eligible to vote in that election. Let us assume, all else equal, that the the vote share in the population for Party A is 24%. This is what we call the **true population parameter.** And our goal is to estimate this parameter with statistical methods, since it is more efficient. Think about Germany, we cannot ask 80 million people before the election to give us their thoughts, so what we do instead, is to draw a representative sample of less citizens from that sample to infer to the population. The problem is that even if the sample is representative, it could be that our sample today gives us a vote share of Party A of 23%, but tomorrow we would get 25% (This could have several reasons, can you think of some?).

In the following, let us assume we draw 1000 samples before the election, to get the vote share of Party A. We know that our true population parameter is 24%. But before we start, we have to set a rule again. This rule is basically the definition of confidence intervals:

-   In 95% of all samples, that could be drawn, the confidence intervals will cover the true population parameter.

So if we draw 1000 samples, in 950 the confidence intervals have to cover the true population parameter:

```{r simulating 100 CIs}
#Since this is a simulation we need to set a seed
set.seed(187)

#We will need to have vectors for the upper confidence interval and the lower one
lower_ci <- numeric(100)
upper_ci <- numeric(100)
estimates <- numeric(100)

#This loop represents 
for(i in 1:length(lower_ci)) {
  
  Y <- rnorm(100, mean = 24, sd = 2)
  estimates[i] <- Y[i]
  lower_ci[i] <- Y[i] - 1.96 * 24 / 10
  upper_ci[i] <- Y[i] + 1.96 * 24 / 10
  
}

#Let us bind both vectors together 
CIs <- data.frame(estimates, lower_ci, upper_ci)

#Print it 
head(CIs)
```

Now we have drawn our 1000 samples and computed our estimates as well as their corresponding intervals. Thus, 1000 samples about the vote share of Party A. Let us check, if the true population parameter is 95% of times within our computed confidence intervals:

```{r plotting simulated CIs}
#Getting the true mean 
true_mean <- 24

#First, we identify those who are not including 24 our true population parameter
CIs$missed <- ifelse(CIs$lower_ci > true_mean | CIs$upper_ci < true_mean, "Out", "In")

#Let us give every sample an identification number 
CIs$id <- 1:nrow(CIs)

#Plotting it 
ggplot(data = CIs) +
  geom_pointrange(
    aes(
      x = estimates, # point value
      xmin = lower_ci, # lower CI
      xmax = upper_ci, # upper CI
      y = id, # y axis - just observation number
      color = missed
    ) # color varies by missed variable
  ) +
  geom_vline(
    aes(xintercept = true_mean), # add vertical line at true_mean
  ) +
  scale_color_manual(values = c("azure4", "red")) + 
  theme_minimal() + 
  labs(
    title = "Confidence Interval for Mean",
    subtitle = "Population mean equals 24",
    x = "Estimates",
    y = "Sample",
    color = "Is true population parameter inside the CI?"
  ) +
  theme(legend.position = "top") + # switch the legend to the top
  scale_x_continuous(breaks = c(seq(15, 30, by = 1)))
```

Well, we can see that the majority of the computed intervals include the true population parameter. But there are three, which are not. That is within our definition.

What is now with confidence intervals for coefficients of linear regression? Well, it is the same story, we compute an estimate, in this case a coefficient. The true population parameter is unknown, but we know that the true population parameter is within 95% of the intervals.

Its calculation is fairly easy and you already saw it:

$$
CI_{lower} = \beta_i - 1.96 * SE(\beta_i) \\\
CI_{upper} = \beta_i + 1.96 * SE(\beta_i)
$$

```{r computing CIs by hand for model 1}
#Let us look at the confindence intervals of model 1 
confint(model1)

#Let us compute the confidence values by hand for the intercept and x 
ci_lower_int <- model1$coefficients[1] - 1.96 * summary(model1)$coef[, "Std. Error"][1]
ci_upper_int <- model1$coefficients[1] + 1.96 * summary(model1)$coef[, "Std. Error"][1]

#Print it 
print(ci_lower_int)
print(ci_upper_int)

#Estimate X 
ci_lower_est <- model1$coefficients[2] - 1.96 * summary(model1)$coef[, "Std. Error"][2]
ci_upper_est <- model1$coefficients[2] + 1.96 * summary(model1)$coef[, "Std. Error"][2]

#Print it
print(ci_lower_est)
print(ci_upper_est)
```

## Multivariate Regression

The world is a complex place and of course if we have a dependent variable Y, let us say for example income, we cannot explain it exclusively by the years of education or exclusively by the profession or any other single factor. Rather it is plausible that all these factors matter. The **Multivariate Linear Regression Model** allows us that we explain the variation in our dependent variable Y with multiple independent variables X. Mathematically, the systematic component changes like this:

$$ Y_i = \beta_0 + \beta_1X_{1i} + \beta_2X_{2i} + ... + \beta_kX_{ki} + \epsilon_i $$

where

-   the index ***i*** runs over the observations (Respondents, Countries,...), *i* = 1,...,*n*

-   $Y_i$ is the dependent variable, the variable we want to explain

-   $X_{1i}$ is the first independent variable or explanatory variable.

-   $X_{2i}$ is the second independent variable or explanatory variable.

-   $X_{ki}$ is the *k*-th independent variable or explanatory variable, where *k* is the number of all our independent variables in our model.

-   $\beta_0$ is the *intercept* of the regression line

-   $\beta_1$ is the *slope* of the regression line of the first explanatory variable.

-   $\beta_2$ is the *slope* of the regression line of the second explanatory variable.

-   $\beta_k$ is the slope of the regression line of the *k*-th explanatory variable, where *k* is the number of all our independent variables in our model.

-   $\epsilon_i$ is the *error term*, thus how our observed data differs from actual population data (e.g. Measurement Error).

In `R`, we can implement a multivariate model very easy with the already known `lm()` function. Let us run a multiple regression with our independent Variable X and our categorical Variable Z, the systematic component for this model looks like this:

$$ Y_i = \beta_0 + \beta_1X_i + \beta_2Z_i + \epsilon_i $$

Let us compute the slopes:

```{r computing model for multivariate analysis}
lm(y ~ x + categorical_variable, data = df) %>% 
  summary()
```

We can see, that this works fine, we only had to add the independent variable in the command such as in our systematic component. Regarding the interpretation it is analogous to the normal linear regression one. For a one-unit increase in our independent variable x, the dependent variable y increases 0.67 units on average, holding all else constant. For category 1 in comparison to category 0, the dependent variable y increases 0.52 units on average, holding all else equal. Now, we can put the calculated coefficients into our systematic component and make predictions:

$$ Y_i = -0.54 + 0.67X_i + 0.52Z_i + \epsilon_i  $$

If we want to visualize this, we would need a three-dimensional coordinate system. Why? Because every independent variable is one-dimension if you want. I said that *k* is the number of our independent variables, technically it is the number of dimensions our model has. I am not a fan of plotting graphs more than three dimensions and I do not recommend it to you either.

### Model Fit: Adjusted R-squared

The last important aspect of Multivariate Regression is the Adjusted R-squared measure. Reconsider, the calculation of the classical R-squared:

-   TSS (Variation in the DV): $TSS =\sum(y_i - \bar{y})^2$ , we just subtract our actual values (`df$y`) from its mean and square it to avoid negative numbers. This gives us the total variation of our dependent variable.

-   ESS (Variation we explain in the DV): $ESS = \sum(\hat{y_i} - \bar{y})^2$ , now we use our predicted values (`df$y_hat`) instead of our actual values. That gives us the variation in the dependent variable, we can explain with our model.

-   $R^2$ (The Variation we can predict from our model): $R^2 = \frac{ESS}{TSS}$ , well to get the proportion we just divide the variation we can explain from our DV from the actual variation through the total variation in the DV. If these two values are the same, thus our model predicts all the variation in our dependent variable and this $R^2$ is 1.

The problem with the classical R-squared is, that if you would add useless independent variables to it, the classical R-squared would decrease, although your model did not increase in explanatory power. This is called **overfitting**. However, adjusted R-squared will account for that problem by introducing a "penalty" for every additional variable. Mathematically, it looks like this:

$$ Adj.R^2 = 1 - \frac{(1 - R^2)*(N - 1)}{N - k - 1} $$

where

-   $R^2$ is our classical R-squared calculated ($\frac{TSS}{ESS}$)

-   $N$ is the number of observations in our sample

-   $k$ is the number of independent variables

In `R`, we can extract the adjusted R-squared simply from our model in chunk, multivariate regression:

```{r calculating adj r2}
#running multivariate model
multivariate_model <- lm(y ~ x + categorical_variable, data = df)

#Getting summary
summary(multivariate_model)

#Extract Adjusted R-squared
summary(multivariate_model)$adj.r.squared

#Calculating by hand
adj_r_squared <- 1 - (((1-summary(multivariate_model)$r.squared) * (nrow(df) - 1))/(nrow(df) - 2 - 1))

#printing it
print(adj_r_squared)
```

### Omitted Variable Bias

I already mentioned one (abstract) reason why we should include other variables in our model. But there is more to it: You could find effects between two variables X and Y, but it could be that in Reality there is not an association. For example, let us say you collect data about ice cream and shark attacks. Ice cream sales is your independent variable and you want to explain the number of shark attacks, here is your data:

```{r omitted variable bias}
# Set seed for reproducibility 
set.seed(0)  

# Number of data points 
n <- 100

# Simulate diet data (assuming a normal distribution) 
temperature <- rnorm(n, mean = 1500, sd = 200)  

# Simulate exercise data (assuming a normal distribution) 
ice_cream_sales <- rnorm(n, mean = 3, sd = 1)  

# Simulate weight loss data 
violence_crime_true <- 0.2 * temperature - 
  0.5 * ice_cream_sales + 
  rnorm(n, mean = 0, sd = 5) 

# Create a data frame 
data <- data.frame(temperature = temperature,ice_cream_sales = ice_cream_sales,                     violence_crime_true = violence_crime_true)  

# Fit a model without including the diet variable 
model_without_temperature <- lm(violence_crime_true ~ ice_cream_sales, data = data)

#Fit a model with only the temperature variable

model_with_only_temperature <- lm(violence_crime_true ~temperature,                        data = data)

# Fit a model including both diet and exercise variables 
model_with_temperature <- lm(violence_crime_true ~ ice_cream_sales + temperature,                        data = data)  

# Output the summary of both models 
summary(model_without_temperature) 
summary(model_with_only_temperature)
summary(model_with_temperature)  

#Let us display both models next to each other
#EDIT: I created this function specifically, the code for the function is at the top.   
table_ovb(model_without_temperature, model_with_temperature)
```

We can see that the coefficient changes dramatically. What happened? Well, one important assumption of linear regression is that the error term captures all variance not explained by our model and is **not correlated** with the independent variable(s) nor the dependent variable. But if there is unexplained variation in our model that is correlated with our independent variable, then this assumption is violated. In our example, we can see that ice cream sales coefficient in the first model is biased, because ice cream sales and is correlated to temperature. The warmer it gets, the more ice cream is sold. But, the warmer it gets, the more violent people get, therefore we have an omitted variable and that is temperature. When we include temperature in the model, we see the problem of omitted variable bias: It biases our coefficients, by either overestimating (like in our example) or by underestimating it. What we should do in such a case, is to delete the omitted variable, which is the drastically changing variable (ice cream sales in our case). This is also the reason, why people talk about additional variables as control variables in a multiple linear model. This way you can control if an association between two variables is due to omitted variable bias or other variables, which can explain the variation better.

### Multicollinearity

Another, and I promise, the last OLS assumption, which has to tested is that **there is no Multicollinearity**. The concept is simple: The independent variables should not be correlated. In our previous example, ice cream sales and temperature were correlated. This would have hurt these assumption. In strong cases, multicollinearity can bias our estimates, so that they gain statistical significance and lead us to wrong conclusions. Let us look at an obvious example. You want to find out how the grades of children is affected by different factors. You choose 2 factors: The time they spent on doing their homework (*learning time*) and the time they spent on playing video games (*gaming time*). The systematic component looks like this:

$$
Grades_i = \beta_0 + \beta_1*\text{learning time}_i + \beta_2 *\text{gaming time}_i + \epsilon_i
$$

Let us compute them:

```{r Multicollinearity}
# Set seed for reproducibility
set.seed(42)

# Number of samples
n <- 100

# True coefficients
beta_0 <- 80
beta_1 <- 1.5
beta_2 <- 1.5

# Generate independent variables
learning_time <- runif(n, 1, 10)
gaming_time <- 0.7 * learning_time + 
  sqrt(1 - 0.7^2) * rnorm(n, sd = 1) 

#generate error term
epsilon <- rnorm(n, 0, 3)

# Generate grades
grades <- beta_0 + beta_1 * learning_time + beta_2 * gaming_time + epsilon

# Create a data frame
df_grades <- data.frame(learning_time,
                   gaming_time,
                   grades)

# Display first few rows of the data frame
grades_model <- lm(grades ~ learning_time + gaming_time, 
                   data = df_grades)

#Getting the summary
summary(grades_model)
```

By looking at the model, we could conclude that the more a student learns, the better its grades on average, holding all else constant. So far, so clear, but the same goes for gaming time. Why is that the case? Because if we think that a student has per day 3 hours, which the student can assign to either learning or gaming, than both are correlated, because assigning 2 hours to learning means 1 hour for gaming, 0.5 hours for learning means 2.5 hours for gaming and so forth. This means both coefficients are explaining each other and bias each other. How to detect them?

#### Testing Correlations to each other

The first technique is to check the correlations of the variables to each other beforehand. You can do that two-ways: Just print out a correlation table:

```{r testing correlations}
#First store the variables you need in a seperate data frame 
cormatrix_data <- df_grades %>% 
  select(learning_time, gaming_time)


#Second, calculate the table, the 2 at the end are the dimensions
cormatrix <- cor(cormatrix_data) #Calculate the correlations
round(cormatrix, 2) #round it to the second digit and display it 

#We could have also done this code in one step 

#df_grades %>% 
#  select(learning_time, gaming_time) %>% 
#  cor() %>% 
#  round(2) 
```

We can see that the correlation between both variables is way to high. Now, with only two variables the table works fine, but what if we have, let us say, 20 variables? It could get messy, therefore you could also use the correlation matrix, which I introduced in the chapter before. In this case, it does not make sense, since it would just print out one block. But keep it nevertheless in mind for the future.

#### Variance of Inflation (VIF)

Another measure for multicolinearity and probably the most famous one, is the Variance-of-Inflation (VIF) factor. Intuitively spoken, this measure fits models with multiple variables, by calculating the variance of each variable. Then it fits a model with only one independent variable and calculates the variance of it. The result is a measure that displays high values if the variance of a variable increases, when other variables are added. That is exactly what this measure does.. The formula of it is really simple:

$$
VIF_i = \frac{1}{1 - R^2_i} 
$$

where, the Variance of inflation (VIF) of variable *i* is calculated by 1 divided by 1 - the R-squared ($R^2_i$) of the regression with only that variable.

In `R`, we can use the `VIF()` function from the `car` - package to do it.

```{r vif}
#We only have to use the function VIF() on our model 
vif(grades_model)
```

The rule is that if a value exceeds 10, it is considered critical. We should always test for multicolinearity, and if we detect it, run the regression separately without the correlated variables.

## Categorical Variables

As I mentioned in the first chapter, there are different types of variables, let us reconsider them:

|               |               |                                              |
|-------------------|-------------------|----------------------------------|
| **Numeric**   | Numbers       | `c(1, 2.4, 3.14, 4)`                         |
| **Character** | Text          | `c("1", "blue", "fun", "monster")`           |
| **Logical**   | True or false | `c(TRUE, FALSE, TRUE, FALSE)`                |
| **Factor**    | Category      | `c("Strongly disagree", "Agree", "Neutral")` |

Dependent variables must be numeric, when conducting linear regression. However, independent variables can take be scaled differently. Numeric variables are the easiest case, the interpretation is as mentioned in the previous examples. But categorical variables are differently to interpret. Let us inspect the categorical variable in our dataset, called `categorical_variable`:

```{r printing table of categorical variable}
table(df$categorical_variable) 
```

As we can see our data set now contains a categorical variable with two categories, named "A" and "B". Those categories could be anything: Female and Male, bought a product or did not buy a product, vaccine or placebo and so on. What happens, when we now run a model?

```{r model with categorical variable}
#running a model with a categorical variable
model2 <- lm(y ~ categorical_variable, 
             data = df) 

#Getting the summary
summary(model2) 
```

We see a coefficient of 0.64 that is fine. But as you see the category "category_A" is not display, why? Categorical variables are calculated based on so-called "reference categories". R determines the reference category based on alphabetical order. In this case the category "A" is the reference category. The reference category is named like this, since the computed coefficients refer to it. The reference category "A" takes on the value 0. The coefficient of the category B is "0.92". That means that category "B" if the a respondent is part of category "B" than the dependent variable Y increases on average 0.64 units in comparison to category "A", holding all else equal.

That sounds technocratic, let us get an intuition. We add a zero to our code and look at the output:

```{r running the model with + 0}
#running the model
model3 <- lm(y ~ categorical_variable + 0, 
             data = df) 
#getting a summary
summary(model3)
```

The coefficients changed, but not really. Let us subtract the coefficient from category A from category B:

```{r subtracting categories to get coefficient}
#results
result <- coefficients(model3)[2] - coefficients(model3)[1] #coefficents can be extracted this way

#printing it
result
```

We get the same coefficient as above and that is what R does automatically when computing coefficients of categorical variables. The reference category is scaled to 0 and the other categories are following.

### Interaction Effects

One technique, which is important and widely used in statistics are interaction effects. To keep things simple concentrate on our independent variable X and our categorical variable Z. As mentioned, Z shows us the coefficients for our categories "A" and "B". But what if we could further investigate the dynamics of the group? Interaction effects allow us to multiply our variable X by our categorical variable Z. Mathematically our systematic component looks like this:

$$
Y_i = \beta_0 + \beta_1X_i + \beta_2Z_i + \beta_3X_i*Z_i + e_i 
$$

where

-   the index *i* runs over the observations (Respondents, Countries,...), *i* = 1,...,*n*

-   $Y_i$ is the dependent variable, the variable we want to explain

-   $X_i$ is the independent variable or explanatory variable.

-   $Z_i$ is our categorical variable

-   $\beta_0$ is the *intercept* of the regression line

-   $\beta_1$ is the *slope* of the regression line

-   $\epsilon_i$ is the *error term*, thus how our observed data differs from actual population data (e.g. Measurement Error).

In the following, I will show you an example how to use interaction effects. Let's say that I conducted a survey among 1000 respondents and asked them about how many hours they spent on R online courses such as this one. Then I asked about their coding ability. Lastly, I asked if they learned with this course or with other courses (0 = other courses, 1 = this course). Now we want to find out if spending more hours on a course lead to a higher coding ability in R. Furthermore, we want to find out if this course in comparison to other courses lead to a higher ability. To do so, we interact hours spent on courses with the variable indicating if the respondent worked through other courses or this course.

-   Interactions effects can be implemented in R, by simply writing it explicitly into the function `lm()`, either with an asterisks `*` or a double point `:`, let us have a look it:

```{r model with interaction effect}
# Setting seed for reproducibility
set.seed(123)

# Generate hours spent on a course
hours_spent <- runif(100, min = 0, max = 10)

# Generate the course dummy (0 = other courses, 1 = this course)
this_course = sample(c(0, 1), n, replace = TRUE)

# Generate y with interaction effect
coding_ability <- 2 + 0.5 * hours_spent + 0 * this_course + 
  1.5 * hours_spent * this_course + rnorm(n)

# Create a data frame
df_int <- data.frame(hours_spent, this_course, coding_ability)

# Fit the interaction model
model_interaction <- lm(coding_ability ~  hours_spent * this_course, data = df_int)

# Summarizing models
summary(model_interaction)
```

Interaction effects on their own are not intuitive. To get an intuition we have to graphically plot it. The `sjPlot` package offers the `plot_model()` command, which automatically plots so called predicted probabilities. It would be too much to go into detail about predicted probabilities. What is more important is, that we get a graph, which shows the effect of of the interaction:

-   We call `plot_model()` and include our model with the interaction effect `model_interaction`

-   Then we have to call the type and set it to `type="int"`, which explicitly plots interaction effects.

```{r plotting interaction effect, message=FALSE}
plot_model(model_interaction, type = "int") +
  scale_x_continuous(breaks = seq(0,10, 1)) + 
  labs(title = "Coding Ability after this Course in Comparison",
       x = "Hours spent", 
       y = "Coding Ability in R") +
  scale_color_manual(
    values = c("red", "blue"),
    labels = c("Other Courses", "This Course")
  ) +
  theme_sjplot() +
  theme(legend.position = "bottom", 
        legend.title = element_blank())
```

On the x-axis, we have our independent variable and on the y-axis we do have our dependent variable. So far, so normal. But we see two lines: One line for all observations in category "Other Courses" (red line) and one line for all observations with "This Course" (blue line). That is exactly what an interaction does. For every point in our independent variable a probability for y is computed in one category, and one line for every observation in the other category.

To interpret it we follow two steps: First, looking at the direction of the lines. We can see that X has a positive effect on Y. Both lines are increasing with higher X-values. Now, the interaction allows us to compare the effect from X on Y for all groups of the categorical variable. In our example that means, that observations in group 1 display a higher effect than observations in group 0. That could be for several reasons.

Let us fill this example with life: We are studying **the effect of study hours (x)** on **exam scores (y)** for two different groups of students: those **who attend a preparatory course (group 1) and those who do not (group 0).**

**Interpretation:**

1.  **Direction of the Lines:**
    -   Both lines (for group 0 and group 1) are increasing, indicating that more hours spent on courses lead to higher coding ability in R, on average.
2.  **Effect Comparison**:
    -   The slope for group 1 ( attended this course) is steeper compared to group 0 (attended to other courses).

    -   This means that for students who attended this course, each additional hour of study has a larger positive impact on their coding ability in R compared to those who did attend another course

By visualizing and analyzing the interaction effect, we can draw meaningful conclusions about how different factors (like attending a preparatory course) modify the relationship between study efforts and performance outcomes.

## Outlook

This chapter was an introduction to hypothesis testing and basic statistical applications. Further, you were introduced in the most popular and important model, linear regression. The chapter showed not only how linear regression in R is computed, but also how to check if effects are robust and how the models fits. Also extensions to linear regression, namely multivariate regression and categorical variable were introduced.

Linear regression is the basis of nearly everything. Advanced modelling of different classes of dependent variables to even machine learning techniques are based on linear regression. This was the reason, I did not show just linear regression, but also how vulnerable the model is. Poor modelling will always lead to poor results, so we have to aim to check how good our model fits our data and thus research interest.

Further Links:

-   A course I can recommend if you want to have a detailed deep dive into statistics is ["Introduction to Econometrics with R"](https://www.econometrics-with-r.org/index.html) by Christoph Hanck, Martin Arnold, Alexander Gerber, and Martin Schmelzer.

## Exercise Section

To understand linear regression, we do not have to load any complicated data. Let us assume, you are a market analyst and your customer is the production company of a series called "Breaking Thrones". The production company wants to know how the viewers of the series judge the final episode. You conduct a survey and ask people on how satisfied they were with the season final and some social demographics. Here is your codebook:

| Variable     | Description                                                                                                                                                    |
|------------------|------------------------------------------------------|
| id           | The id of the respondent                                                                                                                                       |
| satisfaction | The answer to the question "How satisfied were you with the final episode of Breaking Throne?", where 0 is completely dissatisfied and 10 completely satisfied |
| age          | The age of the respondent                                                                                                                                      |
| female       | The Gender of the respondent, where 0 = Male, 1 = Female                                                                                                       |

Let us generate the data:

```{r}
#setting seed for reproduciability
set.seed(123)

#Generating the data
final_BT <- data.frame(
  id = c(1:10), 
  satisfaction = round(rnorm(10, mean = 6, sd = 2.5)), 
  age = round(rnorm(10, mean = 25, sd = 5)), 
  female = rbinom(10, 1, 0.5)
  )

print(final_BT)
```

### Exercise 1: Linear Regression with two variables

You want to know if age has an impact on the satisfaction with the last episode. You want to conduct a linear regression.

a\. Calculate $\beta_0$ and $\beta_1$ by hand

```{r ch6 exercise 1a, eval=FALSE}

```

b\. Calculate $\beta_0$ and $\beta_1$ automatically with R

```{r ch6 exercise 1b, eval=FALSE}

```

c\. Interpret all quantities of your result: Standard Error, t-statistic, p-value, confidence intervals and the $R^2$.

d\. Check for influential outliers

```{r ch6 exercise 1d, eval=FALSE}

```

### Exercise 2: Multivariate Regression

a\. Add the variable `female` to your regression

```{r ch6 exercise 2a, eval=FALSE}

```

b\. Interpret the Output. What has changed? What stays the same?

c\. Make an interaction effect between age and gender and interpret it!

```{r ch6 exercise 2c, eval=FALSE}

```

d\. Plot the interaction and make the plot nice

```{r ch6 exercise 2d, eval=FALSE}

```